---
title: "BIS 625 Project"
author: "Yiziying Chen"
date: "5/3/2019"
output:
  word_document: default
  html_document: default
---

```{r}
data = read.table("hepatitis.data", sep = ",")[,-20]
colnames(data) = c("class", "age", "sex", "steroid", "antiviral", "fatigue", "malaise", "anorexia","liver_big", "liver_firm", "spleen_palpable", "spiders", "ascites", "varices", "bilirubin", "alk_phosphate", "sgot", "albumin", "protime")
for (i in 1:length(colnames(data))){
  levels(data[,i])[levels(data[,i]) == "?"] <-NA
}


data$bilirubin <- as.numeric(as.character(data$bilirubin))
data$protime <- as.numeric(as.character(data$protime))
data$alk_phosphate <- as.numeric(as.character(data$alk_phosphate))
data$sgot <- as.numeric(as.character(data$sgot))
data$albumin <- as.numeric(as.character(data$albumin))
data$age <- as.numeric(as.character(data$age))
data$sex <- as.factor(data$sex)
data$class <- as.factor(data$class)
data$antiviral <- as.factor(data$antiviral)

pairs(data[, c("class", "age", "sgot","albumin", "alk_phosphate", "protime", "bilirubin", "varices")])
```

##contigency table for OR, RR test for association
```{r}
##create contigency tables
cat.attributes = c("sex", "steroid", "antiviral", "fatigue", "malaise", "anorexia","liver_big", "liver_firm", "spleen_palpable", "spiders", "ascites", "varices")

t = list()
table = list()
for(i in 1:length(cat.attributes)){
  t[[i]] = data[complete.cases(data[ , c("class", cat.attributes[i])]),]
  table[[i]] = table(t[[i]][,cat.attributes[i]], t[[i]]$class)
  colnames(table[[i]]) = c("Death", "Live")
}
#label table row&col names
rownames(table[[1]]) = c("M", "F")
for(i in 2:length(cat.attributes)){
 rownames(table[[i]]) = c(paste(cat.attributes[i], "Yes"), "No")
}
```

##Risk difference, Relative Risk and Odds ratio 
###Chi-square test and Exact test for independence
```{r}
library('epiR')
#test for independence
OR = list()
result = list()
IR.CI = list()
OR.CI = list()
indep.test = list()
fisher = list()
for (i in 1:12){
      result[[i]] = epi.2by2(table[[i]], method = "cohort.count", conf.level = 0.95)
  if (all(as.vector(table[[i]]) > 5)) {
      indep.test[[i]] = data.frame(test.stat = result[[i]]$massoc$chisq.strata$test.statistic, p.value = result[[i]]$massoc$chisq.strata$p.value, attri = cat.attributes[i])
  }
  else{
    fisher[[i]] =fisher.test(table[[i]], alternative = 'two.sided')
    indep.test[[i]] = data.frame(est = fisher[[i]]$estimate, ci.lwr = fisher[[i]]$conf.int[1],ci.upr = fisher[[i]]$conf.int[2], p.value = fisher[[i]]$p.value, attri = cat.attributes[i])
    }
  OR[[i]] = result[[i]]$tab
  IR.CI[[i]] = result[[i]]$massoc$RR.strata.wald
  OR.CI[[i]] = data.frame(result[[i]]$massoc$OR.strata.wald, attri = cat.attributes[i])
}

#find insignificant independence
library(pipeR)
library(rlist)
indep.test %>>%
  list.filter(p.value > 0.1) %>>%
  list.mapv(attri)
```

##hypothesis testing--prop test
```{r}
prop.test(sum(as.numeric(data$class)-1), length(data$class), p = 0.5, alternative = "two.sided", conf.level = 0.95)
```


#3-way table; Breslow-Day test the homogeneous associations; mantelhaen test
```{r}
options(scipen = 999)
library(DescTools)
sig.attri = c("fatigue", "malaise",  "spleen_palpable", "spiders", "ascites", "varices")
comb = combn(sig.attri,2)

t2 = list();table2 = list();dp = list();df = list();ftable = list();mantel = list()
mantel.test = list();bd = list();bd.test = list()
for(i in 1:ncol(comb)){
  t2[[i]] = data[complete.cases(data[ , c("class", comb[1,i], comb[2,i])]),]
  table2[[i]] = table(t2[[i]]$class,t2[[i]][,comb[1,i]], t2[[i]][,comb[2,i]])
  #colnames(table2[[i]]) = c("Death", "Live")
  dp[[i]] = as.vector(table2[[i]])
  df[[i]] <- array(dp[[i]], dim = c(2,2,2))
  dimnames(df[[i]]) <- list(Class = c("DIE", "ALIVE"), attri1 = c("YES", "NO"), attri2 = c("YES", "NO"))
  ftable[[i]] = ftable(df[[i]], row.vars = c("attri1", "attri2"), col.vars = "Class")
  mantel[[i]] = mantelhaen.test(df[[i]], correct = F)
  mantel.test[[i]] = c(mantel[[i]]$statistic, p.value = mantel[[i]]$p.value, conf.int.lwr = mantel[[i]]$conf.int[1], conf.int.upr = mantel[[i]]$conf.int[2], est = mantel[[i]]$estimate, attri1 = comb[1,i], attri2 = comb[2,i])
  bd[[i]] = BreslowDayTest(x = table2[[i]], OR = 1)
  bd.test[[i]] = c(test.stat = bd[[i]]$statistic, p.value = bd[[i]]$p.value, attri1 = comb[1,i], attri2= comb[2,i])
}

#check same homogenuous OR among stratum
bd.test%>>%
  list.filter(p.value > 0.1) %>>%
  list.mapv(c(attri1, attri2)) %>>% matrix(ncol = 2, byrow = T)


#check all OR = 1 among statum
mantel.test %>>%
  list.filter(p.value > 0.1 & !is.na(p.value)) %>>%
  list.mapv(c(attri1, attri2)) %>>% matrix(ncol = 2, byrow = T)
```

#residual analysis
```{r}
library(vcd)
death.margin = list(); odds = list(); chi = list();row.resid = list(); pearson.resid = list(); std.resid = list(); resid.result = list()

for(i in 1:12){
# odds ratio of margin table and chisquare test
odds[[i]] = exp(oddsratio(table[[i]])$coefficients)
chi[[i]] = chisq.test(table[[i]])
#row resid
row.resid[[i]] = chi[[i]]$observed - chi[[i]]$expected
#pearson resid
pearson.resid[[i]] = chi[[i]]$residuals
#standardized resid
std.resid[[i]] = chi[[i]]$stdres
##colnames(std.resid[[i]]) = c(paste(comb[1,i], "Yes"), paste(comb[1,i], "No"))
##rownames(std.resid[[i]]) = c("Death_Yes", "Death_No")
}

std.resid
```


##logistic regression--model selection
```{r warning=F}
library(dplyr)
library(MASS)
t.complete = data[complete.cases(data),]

num.complete = data[complete.cases(data$bilirubin, data$alk_phosphate, data$sgot, data$albumin, data$protime),]

cor(num.complete[,c("bilirubin", "alk_phosphate", "sgot", "albumin", "protime")])

shapiro.test(as.numeric(t.complete$class))
#logit link logistic regression
fit = glm(class ~ age + steroid + fatigue + malaise + spleen_palpable + spiders + ascites+ varices + bilirubin + alk_phosphate + sgot + albumin + protime,family = binomial(link = logit), data = t.complete)
summary(fit)
stepAIC(fit, direction = "both")

                #fit.new = glm(class ~ steroid + spiders+varices + sgot + albumin,family = binomial(link = logit), data = t.complete)
fit.new = glm(class ~ steroid + ascites + varices + sgot + protime,family = binomial(link = logit), data = t.complete)
summary(fit.new)
anova(fit.new, fit, test = "LRT")

                #fit.exp <- glm(formula = class ~ steroid + spiders + varices * albumin + sgot, family = binomial(link = logit), data = t.complete)

##check multicollinearity:
pairs(t.complete[,c("class","steroid" ,"ascites","varices" , "sgot", "protime")])

#add interaction term
fit.exp <- glm(formula = class ~ steroid + ascites + varices + sgot * protime, family = binomial(link = logit), data = t.complete)
summary(fit.exp)
anova(fit.new, fit.exp, test = "LRT")


#probit link logistic regression
probit.fit = glm(class ~ age + steroid + fatigue + malaise + spleen_palpable + spiders + ascites+ varices + bilirubin + alk_phosphate + sgot + albumin + protime, family = binomial(link = probit), data = t.complete)
summary(probit.fit)
stepAIC(probit.fit)

probit.new = glm(formula = class ~ age + steroid + ascites + varices + alk_phosphate + sgot + protime, family = binomial(link = probit), data = t.complete)
summary(probit.new)
anova(probit.new, probit.fit, test = "LRT")
#p(Y = death|X )= pnorm(predict(probit.fit))
#p(Y = live|X) = 1-p(Y = death|X)
#OR = p(Y=death|X) / (1-p(Y = death|X))
##check multicollinearity:
pairs(t.complete[,c("age" ,"steroid" , "ascites" , "varices" , "alk_phosphate" , "sgot" , "protime")])

```

##predictive power
```{r}
# ROC
library(pROC)
par(mfrow = c(1,2))
rocplot.fit.new <- roc(class ~ fitted(fit.new), data = t.complete)
plot.roc(rocplot.fit.new, legacy.axes = T, main = "Predictive Power ROC Curve (logit link)")
txt.fit.new <- paste("AUC = ", round(auc(rocplot.fit.new),4))
text(0.5, 0.5, txt.fit.new)
#auc(rocplot)
probit.rocplot.new <- roc(class ~ fitted(probit.new), data = t.complete)
plot.roc(probit.rocplot.new, legacy.axes = T, main = "Predictive Power ROC Curve (probit link)")
probit.txt.new <- paste("AUC = ", round(auc(probit.rocplot.new),4))
text(0.5, 0.5, probit.txt.new)
```

##model checking
```{r}
##residual
fit.new$df.residual
##deviance
fit.new$deviance
##goodness of fit
#p-value for deviance goodness-of-fit test
1-pchisq(fit.new$deviance, fit.new$df.residual)
qchisq(fit.new$deviance, fit.new$df.residual)
## [1] 0.6360822
logit.resids <- cbind(rstandard(fit.new,type="pearson"), residuals(fit.new,type="pearson"), residuals(fit.new,type="deviance"), rstandard(fit.new,type="deviance"))
resids <- as.data.frame(logit.resids)
colnames(resids) = c("std. pearson", "Pearson", "Deviance",  "std. deviance")

par(mfrow = c(2,3))
plot(fit.new$residuals, pch = 16, main = "Fitted Residuals")
abline(h = 0, lty = 2, col = "blue")
plot(resids$`std. pearson` , pch = 16, main = "Standardized Pearson Residuals")
abline(h = 0, lty = 2, col = "blue")
plot(resids$Pearson, pch = 16, main = "Pearson Residuals")
abline(h = 0, lty = 2, col = "blue")
plot(resids$Deviance, pch = 16, main = "Deviance")
abline(h = 0, lty = 2, col = "blue")
plot(resids$`std. deviance`, pch = 16, main = "Standardized Deviance")
abline(h = 0, lty = 2, col = "blue")

#remove outlier and refit model--greater predictive power
##logit model
which.max(abs(fit.new$residuals))

t.new = t.complete[-72,]

fit1 = glm(class ~ steroid + ascites + varices + sgot + protime,family = binomial(link = logit), data = t.new)
rocplot.fit.new1 <- roc(class ~ fitted(fit1), data = t.new)
plot.roc(rocplot.fit.new1, legacy.axes = T, main = "Predictive Power ROC Curve (logit link)")
txt.fit.new1 <- paste("AUC = ", round(auc(rocplot.fit.new1),4))
text(0.5, 0.5, txt.fit.new1)

probit.new$df.residual
##deviance
probit.new$deviance
##goodness of fit
#p-value for deviance goodness-of-fit test
1-pchisq(probit.new$deviance, probit.new$df.residual)

probit.resids <- cbind(rstandard(probit.new,type="pearson"), residuals(probit.new,type="pearson"), residuals(probit.new,type="deviance"), rstandard(probit.new,type="deviance"))
colnames(probit.resids) = c("std. pearson", "Pearson", "Deviance",  "std. deviance")
probit.resids <- as.data.frame(probit.resids)

par(mfrow = c(2,3))
plot(probit.new$residuals, pch = 16, main = "Fitted Residuals")
abline(h = 0, lty = 2, col = "blue")
plot(probit.resids$`std. pearson` , pch = 16, main = "Standardized Pearson Residuals")
abline(h = 0, lty = 2, col = "blue")
plot(probit.resids$Pearson, pch = 16, main = "Pearson Residuals")
abline(h = 0, lty = 2, col = "blue")
plot(probit.resids$Deviance, pch = 16, main = "Deviance")
abline(h = 0, lty = 2, col = "blue")
plot(probit.resids$`std. deviance`, pch = 16, main = "Standardized Deviance")
abline(h = 0, lty = 2, col = "blue")

#remove outlier and refit model--greater predictive power
##probit model
which.max(abs(probit.new$residuals))
order(abs(probit.new$residuals))
probit.new$residuals[c(52,72,55)]

t.new1 = t.complete[-c(52,72,55),]

prob1 = glm(formula = class ~ age + steroid + ascites + varices + alk_phosphate + sgot + protime, family = binomial(link = probit), data = t.new1)
probit.rocplot.new1 <- roc(class ~ fitted(prob1), data = t.new1)
plot.roc(probit.rocplot.new1, legacy.axes = T, main = "Predictive Power ROC Curve (probit link)")
probit.txt.new1 <- paste("AUC = ", round(auc(probit.rocplot.new1),4))
text(0.5, 0.5, probit.txt.new1)

par(mfrow = c(1,2))
plot.roc(rocplot.fit.new1, legacy.axes = T, main = "ROC for Refitted Model(logit link)")
txt.fit.new1 <- paste("AUC = ", round(auc(rocplot.fit.new1),4))
text(0.5, 0.5, txt.fit.new1)
probit.rocplot.new1 <- roc(class ~ fitted(prob1), data = t.new1)
plot.roc(probit.rocplot.new1, legacy.axes = T, main = "ROC for Refitted Model(probit link)")
probit.txt.new1 <- paste("AUC = ", round(auc(probit.rocplot.new1),4))
text(0.5, 0.5, probit.txt.new1)
```

##multinomial analysis
```{r}
#create multiple levels
summary(t.complete$sgot)
summary(t.complete$bilirubin)
summary(t.complete$alk_phosphate)
summary(t.complete$albumin)
for(i in 1:length(t.complete$alk_phosphate)){

  ##categorical sgot
  if (t.complete$sgot[i] <=28){t.complete$cat.sgot[i] = 1}
  else if (t.complete$sgot[i]<=46){t.complete$cat.sgot[i] = 2}
  else if (t.complete$sgot[i]<=65.25){t.complete$cat.sgot[i] = 3}
  else if (t.complete$sgot[i] <= 84){t.complete$cat.sgot[i] = 4}
  ##normal sgot
  if (t.complete$sgot[i] <=40){t.complete$norm.sgot[i] = 0}
  else {t.complete$norm.sgot[i] = 1}

  ##categorical albumin
  if (t.complete$albumin[i] <=3.4){t.complete$cat.albumin[i] = 1}
  else if (t.complete$albumin[i]<=4){t.complete$cat.albumin[i] = 2}
  else if (t.complete$albumin[i]<=4.2){t.complete$cat.albumin[i] = 3}
  else if (t.complete$albumin[i] <= 6.4){t.complete$cat.albumin[i] = 4}
  ##binary albumin
  if (3.4 <= t.complete$albumin[i] | t.complete$albumin[i] <=5.4){t.complete$norm.albumin[i] = 0}
  else {t.complete$norm.albumin[i] = 1}
  
}



#check response distribution
shapiro.test(data$sgot) #non-normal->use glm
shapiro.test(data$albumin) #non-normal->use glm

#check correlation
pairs(t.complete[,c("sgot","steroid","spleen_palpable" ,"antiviral","fatigue","malaise",  "albumin")])
pairs(t.complete[,c("sgot","anorexia","spiders","ascites" ,"varices","albumin")])
pairs(t.complete[,c("bilirubin" ,"albumin" , "protime" , "alk_phosphate" , "sgot")])

library(VGAM)
##multilevel sgot
sgot.fit <- vglm(factor(cat.sgot) ~ bilirubin + alk_phosphate + albumin + protime, family=multinomial(refLevel = "1"), data = t.complete)
summary(sgot.fit)


sgot.fit.cat <- vglm(factor(cat.sgot) ~ steroid + spleen_palpable +antiviral+fatigue+malaise+anorexia+spiders +ascites + varices, family = multinomial(refLevel = "1"), data = t.complete)
summary(sgot.fit.cat)

lrtest(sgot.fit.cat, sgot.fit)

##binary sgot
bin.sgot.fit <- glm(factor(norm.sgot) ~ bilirubin + alk_phosphate + albumin + protime, family=binomial(logit), data = t.complete)
summary(bin.sgot.fit)


bin.sgot.fit.cat <- glm(factor(norm.sgot) ~ steroid + spleen_palpable +antiviral+fatigue+malaise+anorexia+spiders +ascites + varices, family = binomial(logit), data = t.complete)
summary(bin.sgot.fit.cat)

anova(bin.sgot.fit, bin.sgot.fit.cat, test = "LRT")

##multilevel albumin
albumin.fit <- vglm(factor(cat.albumin) ~ bilirubin + sgot + protime, family=multinomial(refLevel = "1"), data = t.complete)
summary(albumin.fit)

albumin.fit.cat <- vglm(factor(cat.sgot) ~ steroid + spleen_palpable +antiviral+fatigue+malaise+anorexia+spiders +ascites + varices, family=multinomial(refLevel = "1"), data = t.complete)
summary(albumin.fit.cat)

lrtest(albumin.fit.cat,albumin.fit)
```

